---
title: "Modelling the impact of a more transmissable variant in Belgium"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(patchwork)
```

There's much to do about the new variants that get picked up everywhere. I applied the previous insights in the stochastic mechanisms behind the pandemic to model how a more transmissible variant would inflence cases in Belgium. This document explains how.

## Model transmission using a negative binomial distribution

Using a negative binomial distribution can account for the observed overdispersion in the number of infections. The initial work of [Blumberg et al (2014)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4214794/) described how the number of cases generated by $s$ original cases, could be described by a negative binomial distribution with a mean of $Rs$ and a dispersion parameter of $ks$. Here $R$ is the effective reproductive number $R_{eff}$. 

In a [paper by Endo et al](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7338915/), $k$ was estimated at about $0.1$ for a consensus $R_0$ of $2.5$. I use their estimates as a guidance throughout this simulation. 

Based on UK data, [Leung et al](https://www.eurosurveillance.org/content/10.2807/1560-7917.ES.2020.26.1.2002106) estimated the UK variant to be 70-80% more infective. I chose to go with 70%, translating to a $R_0$ value of 4.25. 

## Incorporating measures taken to curb spread.

To move from $R_0$ to the effective reproduction rate, I chose to model infection as a binomial process. The key parameter $p$ then gives a probability that a theoretical infection really leads to an infection. So a value $p = 0.6$ means that the measures taken in the population at large, prevent 40% of the infections that would occur if no action was taken. The size parameter of the binomial distribution can then be taken as distributed following the negative binomial I described earlier. 

Based on $R_0$, $R_{eff}$ and the proportion of immune people, one can calculate the parameter for the binomial distribution as:

$$
p = \frac{R_{eff}}{R_0 (1 - p_{immune})}
$$

```{r}
function(r,reff,immune){
  reff / (r * (1 - immune))
}
```

## Estimate of incubation probability.

According to a [meta-analysis by McAloon et al](https://bmjopen.bmj.com/content/10/8/e039652.full), the probability of transmission x days after infection can be modelled by a log-normal distribution with a mean of 1.63 and a standard deviation of 0.5. This allows for creation of a kernel that gives an proxy for the probability of transmission at a given day.

```{r}
inc_kernel <- dlnorm(seq(1,14), mean = 1.63, sd = 0.5)
# Normalization to sum to 1:
inc_kernel <- inc_kernel / sum(inc_kernel)
```

The probabilities used are given in the following graph:

```{r, fig.height=4, fig.width=8, out.width="100%"}
ggplot(data.frame(x=1:14,y=inc_kernel),
       aes(x,y)) +
         geom_area(col = "red",
                   fill = alpha("red", 0.5),lwd = 1) +
  theme_bw() +
  labs(x  = "days after infection", 
       y = "probability of transmission",
       title = "Probability kernel for transmission after infection")
```

## Determining the SEIR model

Actually, it's not the standard SEIR model at all. I use the same reasoning though. I start by taking the number of cases of the previous 2 weeks. For each day, a negative binomial gives a certain number of theoretical infections. These are multiplied with the kernel probabilities and summed to end up with the theoretical new infections on that day. 

These theoretical infections are then corrected:

 - a fraction is considered immune due to previous infection or vaccination. These are substracted.
 - the resulting number is then used as the size parameter for a binomial distribution, which draws the final number of real infections for that day. 
 
This random prediction is done for each day. The number of infections is added to the immune cohort, the window shifts one day, and a new prediction is made. In this way, I simulate what could happen over the coming period.

## Simulating the impact of a new variant.

To simulate what would happen when a new, more transmissable variant enters Belgium, the entire procedure is done for both variants. I make a set of assumptions in this (naive) simulation:

 - immunity after infection is 100%, regardless of which variant you're infected with.
 - in the most optimistic reports, it's estimated that about 30% of the population in Belgium has had the virus somewhere over the past year. Let's roll with that.
 - Current measures in Belgium have the effective reproduction rate $R_{eff}$ floating around 0.95 the past weeks. Given a $R_0$ value of 2.5 and an immunity of about 30%, this translates to a value of 0.54 for the binomial parameter $p$. That would mean we avoid about 46% of possible infections. 
 - About half of the infections are missed in testing. Given an average of 1900 daily confirmed cases over first two weeks of january, I assume about 3800 real cases every day. 
 - We've done about 100,000 vaccinations in Belgium, and I assume an average vaccination speed of 15,000 per day (for now).
 - Let's assume about 5% of infections are the new variant in Belgium.
 
The assumptions are big and rough, and that's because it's sunday and I don't have the time to finetune them. The functions for this simulation are on github.

```{r,message=FALSE,warning=FALSE}
# This loads all the functions used for the simulations here.
source("https://raw.githubusercontent.com/JoFAM/covidBE_analysis/master/functions/simulate_variants.R")
```

## Results

I chose to run 1000 simulations, using the kernel and the assumptions given above. Keep in mind that the band indicate the variation in the simulation, NOT the uncertainty. We assume all parameters fixed here!

```{r}
set.seed(123)

# Simulation
res <- replicate_series(n = 1000,
                        t = 300,
                        prev = seq(3850, 3750, length.out = 14),
                        kernel = inc_kernel,
                        tot_pop = 11e6,
                        vacc = 100000,
                        vacc_speed = 20000,
                        immune = 3000000,
                        p_binom = 0.54) %>%
  mutate(across(c(ll, ul, median),
                function(x) x/2)) %>%
  mutate(t = Sys.Date() + t,
         var = factor(var,
                      levels = c("v1","v2","tot"),
                      labels = c("Old variant: R0 = 2.5",
                                 "New variant: R0 = 4",
                                 "Total")))

# Simulation vaccination speed lower
res2 <- replicate_series(n = 1000,
                        t = 300,
                        prev = seq(3850, 3750, length.out = 14),
                        kernel = inc_kernel,
                        tot_pop = 11e6,
                        vacc = 100000,
                        vacc_speed = 15000,
                        immune = 3000000,
                        p_binom = 0.54,
                        prop_newvar = 0.05) %>%
  mutate(across(c(ll, ul, median),
                function(x) x/2)) %>%
  mutate(t = Sys.Date() + t,
         var = factor(var,
                      levels = c("v1","v2","tot"),
                      labels = c("Old variant: R0 = 2.5",
                                 "New variant: R0 = 4",
                                 "Total")))


```

```{r , fig.height=6, fig.width=8, out.width="100%"}


# Gather data
p1 <- ggplot(res, aes(x = t, fill = var)) +
  geom_ribbon(aes(ymin = ll, ymax = ul),
              alpha = 0.5) +
  geom_line(aes(y = median, color = var), lwd = 1,
            show.legend = FALSE) +
  labs(x = "Date",
       y = "Confirmed cases",
       fill = "Variant") +
  theme_bw() +
  ggtitle("Vaccination speed 20.000 / day \nNo change in policy") +
  scale_x_date(date_breaks = "2 month",
               date_labels = "%b") +
  scale_y_continuous(limits = c(0,10000))

# Gather data
p2 <- ggplot(res2, aes(x = t, fill = var)) +
  geom_ribbon(aes(ymin = ll, ymax = ul),
              alpha = 0.5) +
  geom_line(aes(y = median, color = var), lwd = 1,
            show.legend = FALSE) +
  labs(x = "Date",
       y = "Confirmed cases",
       fill = "Variant") +
  theme_bw() +
  ggtitle("Vaccination speed 15.000 / day \nNo change in policy") +
  scale_x_date(date_breaks = "2 month",
               date_labels = "%b") +
  scale_y_continuous(limits = c(0,10000))

p1 / p2 + plot_layout(guides = "collect")
```
